Evaluating Sound Similarity Metrics for Differentiable, Iterative Sound-Matching

### 2025-09-17 sections/background_related.tex
Updates from Overleaf

Having looked at the current available literature, we note four major areas of weakness in sound-matching. We target the first two issues in this work, while the latter two issues are left for future work.
\begin{enumerate}
    \item \LossSelect: Is there a \textit{best performing} loss function, or is the selection of the loss function dependent on factors such as the sound domain, synthesizer capability and the desired characteristics of the output? The few existing works addressing this issue evaluate their proposed losses using only simplistic sound synthesis methods\todo{cite?}.
    \item \SynthSelect: There is a lack of diversity in the synthesis methods used in sound-matching. The use of different DSP functions for iterative sound-matching and their effects on the outcome or the interactions with the loss function have not been tested.
    \item \PeriodicLoss: The sinusoidal nature of loss function landscapes is a problem that appears frequently\todo{cite?}. In differentiable settings, this problem causes gradient descent updates to local minima. 
    \item \OutDomain: Sound-matching with out-of-domain sounds is an unexplored area, yet a necessary one for practical applications for sound designers.
\end{enumerate}

\subsection{What Approach Should We Take?}
We argue that isolated benchmarking is a necessary precursor to building robust, generalizable sound-matching systems. To address the problems of \LossSelect{} and \SynthSelect{}, we pursue a methodology with 3 key characteristics that do not appear together in other works.

1. Simple differentiable synthesizers using common DSP functions, filters, and oscillators would be beneficial in addressing the \SynthSelect{} problem. Isolating small-scale, low-parameter synthesizers avoids the confounding effects of large neural proxies or embedding models, making it possible to directly analyze the interaction between synthesis structure and loss function.

2. Evaluation of experiments under multiple loss functions: Addressing the \LossSelect{} problem requires that multiple loss functions are tested. A comparison of performance of different loss functions is rare in previous works~\todo{cite}.

3. Parameters optimized directly with gradient descent without additional neural network approximations. This would complement prior neural and proxy-based methods by providing the missing “low-level experiments” that clarify fundamental interactions between DSP functions and similarity measures.

These characteristics would fill a gap in prior work where there is little insight into the comparison of loss functions under different synthesis conditions, and how the two interact.

>>>>

Having looked at the current literature, we identify four major areas of weakness in sound-matching. We target the first two issues in this work, while the latter two issues are left for future work.
\begin{enumerate}
    \item \LossSelect: It remains unclear whether there exists a universally best-performing loss function, 
    or whether performance depends on factors such as the sound domain, synthesizer architecture, 
    and desired output characteristics. Existing studies typically evaluate losses only with simple synthesis 
    setups~\cite{vahidi2023mesostructures}, leaving open questions about their generality. 
    \item \SynthSelect: There is a lack of diversity in the synthesis methods used in sound-matching. The effects of using different isolated DSP functions in iterative sound-matching and how these interact with loss functions remains largely untested.
    \item \PeriodicLoss: The sinusoidal nature of loss function landscapes is a problem that appears frequently\todo{cite and make clearer}. In differentiable settings, this problem causes gradient descent updates to local minima. 
    \item \OutDomain: Sound-matching with out-of-domain sounds is a largely unexplored area, yet a necessary one for practical applications for sound designers.
\end{enumerate}

\subsection{What Approach Should We Take?}
We argue that isolated benchmarking is a necessary precursor to building robust, generalizable sound-matching systems. To address the problems of \LossSelect{} and \SynthSelect{}, we adopt a methodology with three key characteristics that have not previously appeared together.

\begin{enumerate}
    \item Use simple differentiable synthesizers built from common DSP functions (oscillators, filters, envelopes) to address the \SynthSelect{} problem. Restricting to small-scale, low-parameter models avoids the confounding effects of neural proxies or embedding models, enabling direct analysis of how synthesis method interacts with loss functions. 
    \item Evaluate experiments under multiple loss functions to address the \LossSelect{} problem. Prior work has typically tested losses in isolation, making direct comparisons rare~\todo{cite}.  
    \item Optimize synthesizer parameters directly with gradient descent, without neural approximations. This complements prior proxy-based methods by providing the missing “low-level experiments” that clarify fundamental interactions between DSP functions and similarity measures. 
\end{enumerate}

Together, these design choices supply the controlled experimental evidence missing from the literature, 
clarifying how losses and synthesis methods interact in differentiable, iterative settings, 
and offering insights likely to generalize to more complex domains.

### 2025-09-17 sections/methodology_and_analysis.tex
Updates from Overleaf

\section{Experimental Setup}
\label{sec:experiment_setup}

% For clear clarity and alignment with structural norms, we recommend splitting this into three sections: (1) Experimental Setup (including synthesizer programs, loss definitions, optimization procedure); (2) Experimental Results (performance metrics, statistical tests, and rankings); and (3) Discussion (qualitative interpretation, synthesis-specific observations, and practical implications).

  Towards addressing the problems of \LossSelect{} and \SynthSelect{}, we compare the performance of iterative sound-matching experiments, where four differentiable loss functions are applied to four differentiable synthesizers. What we mean by ``performance'' is the similarity of the final output to the target, as measured by P-Loss, MSS, and---most importantly---manual listening tests. \textbf{We hypothesize that the relative performance of loss functions for sound-matching is influenced by the method of synthesis}. If true, this would strengthen the idea that the selection of the loss function and synthesis method is not an optimization problem, but a subjective choice depending on the needs of the sound designer.

@@

An \textit{experiment} is one iteration of sound-matching given a loss function, target sound,
and synthesizer program. When the final iteration is complete, a score is given (automatically with an algorithm or with manual listening) to an experiment that measures its success in replicating the target sound. Consider the following simplified scenario: we have two loss functions, $L_1$ and $L_2$, and a single synthesizer program $P$ with a target parameter set $\theta^*$ where the target sound is $t$, or $P(\theta^*) = t$. We run a large number (in this case, 300 for each loss function) of experiments where we randomly set the initial ($\theta_0$) and target parameters ($\theta^*$) of $P$ and run an iterative sound-matching loop for each loss function, with maximum number of iterations set to 200. When the loop is terminated, we measure the similarity of the final output sound $x$ (where $P(\theta_{199}) = x$) to $t$, or if using P-Loss we compare $\theta_{199}$ to $\theta^*$. Having 300 experiments gives us a distribution of losses to determine whether $L_1$ or $L_2$ led to better performance. This simplified example leads us to our methodology which has four loss functions, four synthesizer programs, and three different evaluation methods of scoring sound similarity.

\subsection{Loss Function Implementation Details}
\label{sec:loss_implementation}
\subsubsection{STFT Losses}
Due to their ubiquity and lower cost of gradient calculation, we use STFT as the basis of two loss functions. As discussed in Section~\ref{sec:matching_types}, the L1 or L2 distances are the most common methods of comparing different spectrograms, however, there are situations where this might lead to shortcomings. An example given by Vahidi \textit{et al.}~\cite{vahidi2023mesostructures} is two chirplets (tones that increase in pitch) starting and ending at the same frequencies. If one of these tones is slightly shifted in time, then the spectrograms will no longer overlap at any point, despite their sonic similarity (similar to the idea that a linear equation $f(x) = ax$ can never overlap with a slightly shifted version of itself, or $f(x) = ax +\epsilon$). In this example, we do not know if this is an issue of the spectrograms, or the L1 measure used for their comparison. In fields such as computer vision, Structural Similarity Index (SSIM)~\cite{wang2004imagesssim,wang2009mean} and Scale-Invariant Mean Squared Error (SIMSE)~\cite{barron2014shapessimse} have been used as alternatives that improved performance in various tasks relative to L1.

@@

\section{Post hoc Analysis of Loss Performance}
\label{sec:results_posthoc}
As discussed in the previous section and summarized in Figure~\ref{fig:posthoc_evaluation}, for each program and evaluation (or scoring) method, there are four sets of 1000 bootstrapped values corresponding to the four loss functions. A post-hoc test is conducted in two stages in order to determine whether there is a difference in performance between loss functions for each program, and if so, which are the best performers. The first stage determines whether there is a difference in group means, with the null hypothesis being that all groups have similar mean ranks. The second stage ranks the loss functions from best to worst using the non-parametric Scott-Knott test (\gls{NPSK})~\cite{tantithamthavorn2017mvt,tantithamthavorn2018optimization}.
The Kruskal-Wallis test is used for the first stage~\cite{kruskal1952use}; this test pools and ranks all evaluation measures for a program, then tests whether the differences in mean rank for all loss function groups is zero. The Kruskal-Wallis test calculates an H statistic that is compared to a chi-square distribution with k-1 degrees of freedom (k is the number of groups, or 4, and degrees of freedom is 3). Using the significance level of 0.05, if the H statistic is greater than the critical value of the chi-square distribution then the null hypothesis is rejected, meaning that at least one group has a mean rank significantly above others~\cite{kruskal1952use}. Table~\ref{tab:kruskal_auto} shows the Kruskal-Wallis results for every program and evaluation method, and shows that significant differences between loss function performance measures (i.e., the bootstrapped distributions of the evaluation for each loss function) exist in all cases except MSS for the \AmpMod{} program.

@@

\section{Takeaways, Practical Implications, and Caveats}
% (Q1) To what extent do automatic evaluation metrics agree with manual listening tests? (Q2) Can novel loss functions utilizing Dynamic Time Warping (DTW) and Scale-Invariant Mean Squared Error (SIMSE) provide advantages over standard loss functions, and if so, when? (Q3) Is an iterative differentiable optimization a viable strategy for design of sound-matching experiments?
The experiment results provide answers regarding the main hypothesis of this work as well as the secondary questions. Based on these answers, we can make practical recommendations for future research.

>>>>

% Section IV titled "Programs, Methodology, and Results" is overloaded and blends experimental setup with results and interpretation. For clear clarity and alignment with structural norms, we recommend splitting this into three sections: (1) Experimental Setup (including synthesizer programs, loss definitions, optimization procedure); (2) Experimental Results (performance metrics, statistical tests, and rankings); and (3) Discussion (qualitative interpretation, synthesis-specific observations, and practical implications).

\section{Experimental Setup}
\label{sec:experiment_setup}
  We want to compare the performance of iterative sound-matching experiments, where different differentiable loss functions are applied to different differentiable synthesizers. What we mean by ``performance'' is the similarity of the final output to the target, as measured by P-Loss, MSS, and---most importantly---manual listening tests. The results here can provide answers for our stated hypothesis and secondary questions\todo{reference}. \todo{Do I remove?}If our hypothesis is true, this would strengthen the idea that the selection of the loss function and synthesis method is not an optimization problem, but a subjective choice depending on the needs of the sound designer.

@@

What we call an \textit{experiment} is one iteration of sound-matching given a loss function, target sound,
and synthesizer program. When the final iteration is complete, a score is given (automatically with an algorithm or with manual listening) to an experiment that measures its success in replicating the target sound. Consider the following simplified scenario: we have two loss functions, $L_1$ and $L_2$, and a single synthesizer program $P$ with a target parameter set $\theta^*$ where the target sound is $t$, or $P(\theta^*) = t$. We run a large number (in this case, 300 for each loss function) of experiments where we randomly set the initial ($\theta_0$) and target parameters ($\theta^*$) of $P$ and run an iterative sound-matching loop for each loss function, with maximum number of iterations set to 200. When the loop is terminated, we measure the similarity of the final output sound $x$ (where $P(\theta_{199}) = x$) to $t$, or if using P-Loss we compare $\theta_{199}$ to $\theta^*$. Having 300 experiments gives us a distribution of losses to determine whether $L_1$ or $L_2$ led to better performance. This simplified example leads us to our methodology which has four loss functions, four synthesizer programs, and three different evaluation methods of scoring sound similarity.

\subsection{Loss Function Implementation Details}
\label{sec:loss_implementation}
\subsubsection{STFT Losses}
Due to their ubiquity and lower cost of gradient calculation, we use STFT as the basis of two loss functions.  L1 or L2 distances are the most common methods of comparing different spectrograms~\cite{turian2020sorry,richard2025model}, however, there are situations where this might lead to shortcomings. \todo{(this is not addressed by SIMSE so why is it here?) An example given by Vahidi \textit{et al.}~\cite{vahidi2023mesostructures} is two chirplets (tones that increase in pitch) starting and ending at the same frequencies. If one of these tones is slightly shifted in time, then the spectrograms will no longer overlap at any point, despite their sonic similarity (similar to the idea that a 2D plot of say, $f(x) = ax$, can never overlap with a slightly vertically shifted version of itself, or $f(x) = ax +\epsilon$). In this example, we do not know if this is an issue of the spectrograms, or the L1 measure used for their comparison. In fields such as computer vision, Structural Similarity Index (SSIM)~\cite{wang2004imagesssim,wang2009mean} and Scale-Invariant Mean Squared Error (SIMSE)~\cite{barron2014shapessimse} have been used as alternatives that improved performance in various tasks relative to L1.}

@@

% \section{Post hoc Analysis of Loss Performance}
\label{sec:results_posthoc}
As discussed in the previous section and summarized in Figure~\ref{fig:posthoc_evaluation}, for each program and evaluation (or scoring) method, there are four sets of 1000 bootstrapped values corresponding to the four loss functions. A post-hoc test is conducted in two stages in order to determine whether there is a difference in performance between loss functions for each program, and if so, which are the best performers. The first stage determines whether there is a difference in group means, with the null hypothesis being that all groups have similar mean ranks. The second stage ranks the loss functions from best to worst using the non-parametric Scott-Knott test (\gls{NPSK})~\cite{tantithamthavorn2017mvt,tantithamthavorn2018optimization}.
The Kruskal-Wallis test is used for the first stage~\cite{kruskal1952use}; this test pools and ranks all evaluation measures for a program, then tests whether the differences in mean rank for all loss function groups is zero. The Kruskal-Wallis test calculates an H statistic that is compared to a chi-square distribution with k-1 degrees of freedom (k is the number of groups, or 4, and degrees of freedom is 3). Using the significance level of 0.05, if the H statistic is greater than the critical value of the chi-square distribution then the null hypothesis is rejected, meaning that at least one group has a mean rank significantly above others~\cite{kruskal1952use}. Table~\ref{tab:kruskal_auto} shows the Kruskal-Wallis results for every program and evaluation method, and shows that significant differences between loss function performance measures (i.e., the bootstrapped distributions of the evaluation for each loss function) exist in all cases except MSS for the \AmpMod{} program.

@@

% \section{Takeaways, Practical Implications, and Caveats}
% (Q1) To what extent do automatic evaluation metrics agree with manual listening tests? (Q2) Can novel loss functions utilizing Dynamic Time Warping (DTW) and Scale-Invariant Mean Squared Error (SIMSE) provide advantages over standard loss functions, and if so, when? (Q3) Is an iterative differentiable optimization a viable strategy for design of sound-matching experiments?
The experiment results provide answers regarding the main hypothesis of this work as well as the secondary questions. Based on these answers, we can make practical recommendations for future research.
